---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis-CI Build Status](https://travis-ci.org/cboettig/piggyback.svg?branch=master)](https://travis-ci.org/cboettig/piggyback)
[![Coverage status](https://codecov.io/gh/cboettig/piggyback/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/piggyback?branch=master)


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# piggyback

`piggyback` is basically a poor soul's [git LFS](https://git-lfs.github.com/).  GitHub rejects commits containing files larger than 50 Mb.  `git LFS` is not only expensive, it also [breaks GitHub's collaborative model](https://medium.com/@megastep/github-s-large-file-storage-is-no-panacea-for-open-source-quite-the-opposite-12c0e16a9a91). (Someone wants to submit a PR with a simple edit to your docs, they cannot fork)
Unlike git LFS, `piggyback` doesn't take over your standard `git` client, it just perches comfortably on the shoulders of your existing GitHub API.  Data can be versioned by `piggyback`, but relative to `git LFS` versioning is less strict: uploads can be set as a new version or allowed to overwrite previously uploaded data.  `piggyback` works with both public and private repositories.  

`piggyback` is inspired by [datastorr](https://github.com/ropenscilabs/datastorr) from [\@richfitz](http://github.com/richfitz). While `datastorr` provides your data in the R "data package" model of `storr` (your data lives inside an R package that you install and is automatically loaded into memory), `piggyback` tries to imitate the LFS model where your data files are simply stored as part of your repository.  This implementation uses R wrappers to the GitHub API, but an identical client could be implemented in any language with bindings to the GitHub API. 

`piggyback` is designed around two possible workflows: (1) A high-level workflow that parallels `git lfs`, and (2) a simple interface for directly uploading and downloading individual data files to a GitHub release.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/piggyback")
```

## Authentication

No authentication is required to download data from *public* GitHub repositories using `piggyback`. To upload data to any repository, or to download data from *private* repositories, you will need to authenticate first.  To do so, add your [GitHub Token]() to an environmental variable, e.g. in a `~/.Renviron` file in your home directory (or some place private you won't upload), or simply set
it in the R console using:

```r
Sys.setenv(GITHUB_TOKEN="xxxxxx")
```

Try to avoid writing `Sys.setenv()` in scripts -- remember, the goal here is to avoid writing your private token in any file that might be shared, even privately.  

## Downloading data

Download the latest version or a specific version of the data:

```{r}
library(piggyback)
pb_download("cboettig/piggyback", "mtcars.tsv.gz")
```

Or a specific version:

```{r}
pb_download("cboettig/piggyback", "mtcars.tsv.gz", tag = "v0.0.4")
```

Or simply omit the file name to download all assets connected with a given release.  you can also always specify a destination directory to download. 

```{r}
dir.create("data")
pb_download("cboettig/piggyback", dest="data/")
```  


## Uploading data

If your GitHub repository doesn't have any [releases](https://help.github.com/articles/creating-releases/) yet, `piggyback` will help you quickly create one.  Create new releases to manage multiple versions of a given data file. While you can create releases as often as you like, making a new release is by no means necessary each time you upload a file.  If maintaining old versions of the data is not useful, you can stick with a single release and upload all of your data there.  

```{r}
gh_new_release("cboettig/piggyback", "v0.0.1")
```

Once we have at least one release available, we are ready to upload.  By default, `pb_upload` will attach data to the latest release.  

```{r eval=FALSE}
## We'll need some example data first.
## Pro tip: compress your tabular data to save space & speed upload/downloads
readr::write_tsv(mtcars, "mtcars.tsv.gz")

pb_upload("cboettig/piggyback", "mtcars.tsv.gz")
```

You can also simply overwrite the a previous version of the file on an existing release, rather than creating a new release every time:

```{r}
pb_upload("cboettig/piggyback", "mtcars.tsv.gz")
```

This is useful in scripts that may automatically upload their results, or whenever a previous version of a data file is disposable.  


## git-style

For an even simpler way to sync data files to GitHub, `piggyback` provides LFS-like `push` and `pull` methods.  

```{r}
# Not implemented yet
pb_track("*.tsv.gz")

pb_pull()
pb_push()
```

This assumes we are working in a directory that is part of the relevant GitHub repository (i.e. the `repo` field is detected based on the `git` remote address for the current working directory.)  By default this uses the latest available release of the currently active GitHub repository. Specify a particular data release tag with the optional argument, `tag`. Data files can be specified by giving a path to a directory where data files are stored on the repo.  Alternately, file paths or types can be specified in a configuration file, see `pb_watch()`.  Identical files will not be transferred. 

*Developer note*: GitHub API does not report hashes for files.  We will have to upload a metadata file to the release that provides these hashes.  Also could consider using git commit hook with these?


## A Note on GitHub Releases vs Data Archiving

`piggyback` is not intended as a data archiving solution.  Importantly, bare in mind that there is nothing special about multiple "versions" in releases, as far as data assets uploaded by `piggyback` are concerned.  The data files `piggyback` attaches to a Release can be deleted or modified at any time -- creating a new release to store data assets is the functional equivalent of just creating new directories `v0.1`, `v0.2` to store your data.  (GitHub Releases are always pinned to a particular `git` tag, so the code/git-managed contents associated with repo are more immutable, but remember our data assets just piggyback on top of the repo).  

Permanent, published data should always be archived in a proper data repository with a DOI, such as [zenodo.org](https://zenodo.org). Zenodo can freely archive public research data files up to 50 GB in size, and data is strictly versioned (once released, a DOI always refers to the same version of the data, new releases are given new DOIs). `piggyback` is meant only to lower the friction of working with data during the research process.  (e.g. provide data accessible to collaborators or continuus integration systems during research process, including for private repositories.)

